Do expert pianists adapt their pedagogical demonstrations to novices' skills?

### 2023-01-17 paper/manuscript.Rmd
add some figures for supplementary materials
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

classoption       : "man"
output            : papaja::apa6_pdf
---

@@

To address this question, we investigated whether and how experts can exaggerate specific parameters of the action to be acquired depending on learners' skill level. We chose to study expressive music performance, in which experts need to modulate multiple performance parameters such as articulation and dynamics to implement expression [@cancino-chacon_2018]. Musicians can create their unique expression and interpretation of music even if they play exactly the same piece because when and how they change multiple performance parameters are different. By studying expressive performance, we can investigate how precisely experts can highlight specific parameters in a multidimentional space of expression (e.g., tempo, articulation dynamics) depending on the goal of teaching (e.g., to teach articulation or dynamics) even when they are performing the same piece.

@@

The experiment was programmed in Max/MSP (8.1.11; https://cycling74.com/products/max) on a Mac Book Pro with Mac OS X Catalina 10.15.7. A weighted Yamaha MIDI digital piano was used to record participants' performances. The pitch, onset and offset time of each note, and key velocity profiles were obtained from MIDI data using Max/MSP patchers. All auditory feedback was given to participants through headphones (Audio-Technica ATH-M50X). Sheet music was displayed on a computer monitor in front of the participants.

The recorded performances serving as stimuli were all of one piece of music that had already been used in our previous study [@tominaga_2022]. The piece was an excerpt from Clementi's Sonatina Op.36 (No.3) in C major. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points (i.e., the number of notes, the number of intervals between notes) for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter to be played with the right hand only. Two expressive notations (i.e., articulation and dynamics) were added, as shown in *Fig \ref{fig:stim}*.

We created artificial recordings of novices to manipulate the displayed proficiency with regard to the two relevant techniques. There were four types of recordings: 1) recordings where both articulation and dynamics were implemented (both), 2) recordings where only articulation was implemented whereas dynamics was missing (art only), 3) recordings where only dynamics was implemented whereas articulation was missing (dyn only), and 4) recordings where neither articulation nor dynamics was implemented (none). We generated 4 instances for each type, therefore there were 16 stimuli (i.e., recordings) in total. How we generated the recordings is described in *Supplementary Material*.

@@

After participants completed the 16 trials, they were asked to perform the piece with the same instruction as in their baseline performance. At the end of the experiment, participants filled in a questionnaire asking about their demographic information and experience in piano performance/teaching. All participants were fully debriefed at the end of the experiment and informed that the recordings had been artificially generated.

@@

@@

Artificial recordings were generated from performance data that we obtained in our previous study [@tominaga_2022]. We used the data from Experiment 2. In Experiment 2, participants were asked to memorise one piece of music (*Fig 5* (A), @tominaga_2022) without any expressive notation before coming to the experiment. First, we asked participants to perform the memorised piece before starting the experiment to measure their baseline performance. After measuring the baseline performance, participants were asked to perform the piece with either articulation (*Fig 5* (B), @tominaga_2022) or dynamics (*Fig 5* (C), @tominaga_2022) in two conditions (teaching vs. performing). Here, we only used the data from the performing condition to make sure that the performances that we used did not have any particular modulations because of teaching intentions. Therefore, there were two types of performance data (articulation performance: performances played with notated articulation, dynamics performance: performances played with notated dynamics). We obtained 31 valid performances (i.g., performances without pitch errors) as baseline performances, 137 valid performances as articulation performances and 139 valid performances as dynamics performances.

@@

We aimed to generate four types of recordings and each type (i.e., both, art only, dyn only, none) had four instances (see *Apparatus and stimuli*) by using the performance data from @tominaga_2022. First, 16 instances for each type were generated. Second, for each type, we chose the four best instances, which exhibited the characteristic of each type well.

@@

In order to create recordings, we needed information about 1) the pitch order, 2) when each note was pressed (related to tempo), 3) when each note was released (related to articulation) and 4) how fast each note was hit (related to dynamics). Pitch information was already determined and the same for all structures because of the sheet music (*Fig \ref{fig:stim}*).

First, we created 16 basic structures of recordings from the baseline performances. To determine the temporal feature of each structure (i.e., when each key was pressed), three performances out of the 31 valid baseline performances were randomly selected and interonset intervals (IOIs) of the three performances were averaged. Due to the size of the dataset (i.e., 31 performances), some baseline performances were used twice. We made sure that the identical performance was not selected within each random selection (i.e., each selection consisted of three different performances). Thus, we created 16 basic structures for tempi.

Next, we were going to follow the same procedure to determine the sound duration (i.e., when each key was released) of each note (related to articulation) and the key velocity profile (i.e., how fast each key was hit) of each note (related to dynamics). For the sound duration, however, we found that most participants implemented articulation (particularly legato) even in their baseline performances, where they should not implement any particular articulation or dynamics. Therefore, we decided to use the data from only two participants who did not show such tendencies. We obtained four performances from the two participants and used these four performances four times each without performing any averaging so as to have 16 instances in total. For the key velocity profiles, we followed the same procedure as we determined tempi and three performances out of the 31 valid baseline performances were randomly selected and the key velocity profiles of the three performances were averaged. This is how we created 16 instances for three performance features (i.e., tempo, articulation and dynamics). We generated the 16 recordings where neither articulation nor dynamics was implemented (i.e., none) by combining the three performance features made from the baseline performances.

In order to have other three types of recordings (i.e., both, art only, dyn only), we needed to create 16 instances for the sound duration when articulation was implemented and 16 instances for the key velocity profiles when dynamics was implemented. We followed the same procedure as we did to generate none recordings but we used articulation performances (not baseline performances) for the sound duration and dynamics performances (not baseline performances) for the key velocity profiles. To create art only recordings,

@@

After we generated 16 recordings for each type, we asked 3 musician raters including the first author to listen to each generated recording and rate to what extent each expression was implemented between 0 (not at all) and 5 (fully). Based on these ratings, we selected four recordings for each type.

>>>>

classoption       : "man"
output            : papaja::apa6_docx
---

@@

To address this question, we investigated whether and how experts can exaggerate specific parameters of the action to be acquired depending on learners' skill levels. We chose to study expressive music performance, in which experts need to modulate multiple performance parameters such as articulation and dynamics to implement expression [@cancino-chacon_2018]. Musicians can create their unique expression and interpretation of music even if they play exactly the same piece because when and how they change multiple performance parameters are different. By studying expressive performance, we can investigate how precisely experts can highlight specific parameters in a multidimensional space of expression (e.g., tempo, articulation dynamics) depending on the goal of teaching (e.g., to teach articulation or dynamics) even when they are performing the same piece.

@@

The experiment was programmed in Max/MSP (8.1.11; https://cycling74.com/products/max) on a MacBook Pro with Mac OS X Catalina 10.15.7. A weighted Yamaha MIDI digital piano was used to record participants' performances. The pitch, onset and offset time of each note, and key velocity profiles were obtained from MIDI data using Max/MSP patchers. All auditory feedback was given to participants through headphones (Audio-Technica ATH-M50X). Sheet music was displayed on a computer monitor in front of the participants.

The recorded performances serving as stimuli were all of one piece of music that had already been used in our previous study [@tominaga_2022]. The piece was an excerpt from Clementi's Sonatina Op.36 (No.3) in C major. The first 12 measures of the original piece were used and modified so that the piece had an almost equal number of data points (i.e., the number of notes, the number of intervals between notes) for each dependent variable. The modified piece consisted of a 12-measure isochronous melody notated in 4/4 meter to be played with the right hand only. Two expressive notations (i.e., articulation and dynamics) were added, as shown in *Fig \ref{fig:stim}*. The fingering was assigned but participants were told that they did not have to follow the fingering indicated if they were not comfortable with it.

We created artificial recordings of novices to manipulate the displayed proficiency with regard to the two relevant techniques. There were four types of recordings: 1) recordings where both articulation and dynamics were implemented (both), 2) recordings where only articulation was implemented whereas dynamics was missing (art-only), 3) recordings where only dynamics was implemented whereas articulation was missing (dyn-only), and 4) recordings where neither articulation nor dynamics was implemented (none). We generated four instances for each type, therefore there were 16 stimuli (i.e., recordings) in total. How we generated the recordings is described in *Supplementary Material*.

@@

After participants completed the 16 trials, they were asked to perform the piece with the same instruction as in their baseline performances. At the end of the experiment, participants filled in a questionnaire asking about their demographic information and experience in piano performance/teaching. All participants were fully debriefed at the end of the experiment and informed that the recordings had been artificially generated.

@@

There are possibilities that it may be difficult in general for experts to flexibly adjust to different action parameters while teaching or they might have chosen to focus on the most salient parameters or the ones that they find most relevant. For example, when teaching dance, it is considered that good teachers identify and build the basic knowledge and skills first and then add complexity progressively from the basic foundation [@mainwaring_2010]. When teaching social dance, a teacher started with a basic step as the hustle and then moved onto more complex steps progressively like the waltz and the cha cha [@lobel_2021]. Karl Leimer, a German music teacher and pianist, wrote in his booked that students should perform with legato at first in finger exercises such as the playing of scales [@gieseking_1972]. In both dance and music examples, the ultimate goal of skill acquisition is that students are able to use multiple techniques flexibly while performing. However, it seems that students need to build complex skills progressively based on the basics during the learning process. Therefore, identifying and focusing on the most relevant problem for each student may be more effective to facilitate skill acquisition.

@@

Artificial recordings were generated from performance data that we obtained in our previous study [@tominaga_2022]. We used the data from Experiment 2. In Experiment 2, participants were asked to memorise one piece of music (*Fig 5* (A), @tominaga_2022) without any expressive notation before coming to the lab. First, we asked participants to perform the memorised piece before starting the experiment to measure their baseline performances. After measuring the baseline performance, participants were asked to perform the piece with either articulation (*Fig 5* (B), @tominaga_2022) or dynamics (*Fig 5* (C), @tominaga_2022) in two conditions (teaching vs. performing). Here, we only used the data from the performing condition to make sure that the performances that we used did not have any particular modulations because of teaching intentions. Therefore, there were two types of performance data (articulation performance: performances played with notated articulation, dynamics performance: performances played with notated dynamics). We obtained 31 valid performances (i.g., performances without pitch errors) as baseline performances, 137 valid performances as articulation performances and 139 valid performances as dynamics performances.

@@

We aimed to generate four types of recordings (i.e., both, art-only, dyn-only, none) and each type had four instances (see *Apparatus and stimuli*) by using the performance data from @tominaga_2022. First, 16 instances for each type were generated. Second, for each type, we chose the four best instances, which exhibited the characteristic of each type well.

@@

In order to create recordings, we needed information about 1) the pitch order (the order of each key press), 2) the onset time of each key press (related to tempo), 3) the offset time of each key press (related to articulation) and 4) the key velocity of each key press (related to dynamics). Pitch information was already determined and the same for all instances based on sheet music (*Fig \ref{fig:stim}*).

First, we created 16 temporal structures of recordings from the baseline performances. Three performances out of the 31 valid baseline performances were randomly selected and interonset intervals (IOIs) of the three performances were averaged for each interval. Due to the dataset size (i.e., 31 performances), some baseline performances were used twice. We made sure that the identical performance was not selected within each random selection of three performances.

Next, we were going to follow the same procedure to determine the sound duration of each note (related to articulation) and the key velocity profile of each note (related to dynamics). For the sound duration, however, we found that most participants implemented articulation (particularly legato) even in their baseline performances, where they should not implement any particular articulation or dynamics. Therefore, we decided to use the data from only two participants who did not show such tendencies. We obtained four performances from the two participants and used these four times each without performing any averaging so as to have 16 instances in total. For the key velocity profiles, we followed the same procedure as we determined tempi and three performances out of the 31 valid baseline performances were randomly selected and the key velocity profiles of the three performances were averaged for each key. This is how we created 16 instances for three performance features (i.e., tempo, articulation and dynamics). We generated the 16 recordings where neither articulation nor dynamics was implemented (i.e., none) by combining the three performance features made from the baseline performances.

In order to have other three types of recordings (i.e., both, art-only, dyn-only), we needed to create 16 instances for the sound duration when articulation was implemented and 16 instances for the key velocity profiles when dynamics was implemented. We followed the same procedure of averaging as we did to generate none recordings but we used articulation performances to determine the sound duration and dynamics performances to determine the key velocity profiles. To create art-only recordings, we replaced the articulation performance feature of none recordings with the one generated from articulation performances. Similarly for dyn-only recordings, we replaced the dynamics performance feature of none recordings with the one generated from dynamics performances. Lastly, to create both recordings, we replaced the articulation and dynamics performance features of none recordings with the ones generated from articulation and dynamics performances (*Fig \ref{fig:supplementary}*). Thus, we generated 16 recordings for each type.

@@

After we generated 16 recordings for each type, we asked 3 musician raters including the first author to listen to each generated recording while sheet music (*Fig \ref{fig:stim}*) was displayed and rate to what extent each notated expression (i.e., articulation, dynamics) was implemented between 0 (not at all) and 5 (fully). Based on these ratings, we selected four recordings, which represented each type of recordings the most. In order to fix the temporal feature of performance for four selected instances per type, we randomly selected 4 tempo performance features from 16 instances and then replaced the tempo performance features of each recording type with the ones we selected.

### 4. Tweaking selected recordings

For dyn-only and none recordings, the recordings sounded too much staccato. We calculated the difference between the mean of each recording for dyn-only and none recordings and the grand mean of art-only and both recordings in terms of KOT and add that difference to all KOT values for dyn_only and low, so that the KOT values should shift closer to the grand mean of art_only and high stimuli.

```{r supplementary, out.width = "100%", fig.cap = "\\label{fig:supplementary}Schematic representation of stimuli generation processes", echo = FALSE}
here::i_am("README.md")
stim <- image_border(image_read(here("paper/image", "supplementary.png")), "#FFFFFF", "140x140")
image_write(stim, path = here("paper/image", "Fig10.tiff"), format = "tiff")
stim
```

```{r analysis-supplementary, include = FALSE}
here::i_am("README.md")
source(here("material/expstimuli/audio/tweaked", "analysis.R"), chdir = TRUE)

ioi_profiles <- gghistogram(dt_ioi[SubArt != "NA"], x = "IOI", facet.by = "MidFile", bins = 50, rug = TRUE)

kot_profiles <- gghistogram(dt_kot[SubArt != "NA"], x = "FixedKOT", color = "SubArt", facet.by = "MidFile", bins = 50, rug = TRUE)

vel_profiles <- gghistogram(dt_vel[SubArt != "NA"], x = "Velocity", facet.by = "MidFile", color = "SubDyn", bins = 50, rug = TRUE)

vel_diff_profiles <- gghistogram(dt_vel_diff[SubDyn == "FtoP" | SubDyn == "PtoF"], x = "Diff", facet.by = "MidFile", color = "SubDyn", bins = 50, rug = TRUE)

# save plots
ggsave(here("paper/image", "s_ioi_profiles.png"), plot = ioi_profiles, dpi = 300, width = 7, height = 5)
img_s_ioi_profiles <- image_read(here("paper/image", "s_ioi_profiles.png"), density = 300)

ggsave(here("paper/image", "s_kot_profiles.png"), plot = kot_profiles, dpi = 300, width = 7, height = 5)
img_s_kot_profiles <- image_read(here("paper/image", "s_kot_profiles.png"), density = 300)

ggsave(here("paper/image", "s_vel_profiles.png"), plot = vel_profiles, dpi = 300, width = 7, height = 5)
img_s_vel_profiles <- image_read(here("paper/image", "s_vel_profiles.png"), density = 300)

ggsave(here("paper/image", "s_vel_diff_profiles.png"), plot = vel_diff_profiles, dpi = 300, width = 7, height = 5)
img_s_vel_diff_profiles <- image_read(here("paper/image", "s_vel_diff_profiles.png"), density = 300)
```

```{r plot-s-ioi, out.width = "100%", fig.cap = "\\label{fig:s-ioi}A"}
img_s_ioi_profiles
```

```{r plot-s-kot, out.width = "100%", fig.cap = "\\label{fig:s-kot}A"}
img_s_kot_profiles
```

```{r plot-s-vel, out.width = "100%", fig.cap = "\\label{fig:s-vel}A"}
img_s_vel_profiles
```

```{r plot-s-vel-diff, out.width = "100%", fig.cap = "\\label{fig:s-vel-diff}A"}
img_s_vel_diff_profiles
```

### 2023-01-17 paper/manuscript.tex
add some figures for supplementary materials
Signed-off-by: Atsuko Tominaga <Tominaga_Atsuko@phd.ceu.edu>

There are possibilities that it may be difficult in general for experts to flexibly adjust to different action parameters while teaching or they might have chosen to focus on the most salient parameters or the ones that they find most relevant. For example, when teaching dance, it is considered that good teachers need to identify and build the basic knowledge and skills first and then add complexity progressively from the basic foundation (Mainwaring \& Krasnow, 2010). For example, when teaching social dance, a teacher started with a basic step as the hustle and then moved onto more complex steps progressively like the waltz and the cha cha (Lobel, 2021). Karl Leimer, a German music teacher and pianist, wrote in his booked that when doing finger exercises, students should perform with legato at first (Gieseking \& Leimer, 1972). Legato creates round and mellow tones, which is one of the most essential elements of piano touches in performance. In both dance and music examples, the ultimate goal of skill acquisition is that students are able to use multiple techniques flexibly while performing. However, it seems that students need to build complex skills based on basic ones during the learning process. Therefore, identifying and focusing on the most relevant problem for each student may be more effective to facilitate learning.

>>>>

There are possibilities that it may be difficult in general for experts to flexibly adjust to different action parameters while teaching or they might have chosen to focus on the most salient parameters or the ones that they find most relevant. For example, when teaching dance, it is considered that good teachers identify and build the basic knowledge and skills first and then add complexity progressively from the basic foundation (Mainwaring \& Krasnow, 2010). When teaching social dance, a teacher started with a basic step as the hustle and then moved onto more complex steps progressively like the waltz and the cha cha (Lobel, 2021). Karl Leimer, a German music teacher and pianist, wrote in his booked that students should perform with legato at first in finger exercises such as the playing of scales (Gieseking \& Leimer, 1972). In both dance and music examples, the ultimate goal of skill acquisition is that students are able to use multiple techniques flexibly while performing. However, it seems that students need to build complex skills progressively based on the basics during the learning process. Therefore, identifying and focusing on the most relevant problem for each student may be more effective to facilitate skill acquisition.
